Disturbance-aware minimum-time planning strategies for motorsport vehicles with probabilistic safety certificates

### 2025-04-15 stochastic_vehicle_dynamics.tex
[MGa] small mod. No H steps ahead (open-loop) back-off

Accordingly, the continuous-time stochastic trajectory planning can be framed as the following nonlinear optimal control problem
\begin{align} 
& \underset{\bmu(t), \bu(t), \bP(t)}{\text{minimize}} \quad J(\bmu(t), \bu(t), \bP(t)) \label{eq:OCPcost} \\
& \text{s.t.} \quad \dbmu(t)           = f(\bmu(t), \bu(t)) \label{eq:OCPdyn} \\
& \phantom{\text{s.t.} \quad} \bmu(0)  = \bmu_0 \label{eq:OCPdynIC} \\
& \phantom{\text{s.t.} \quad} \dbP(t) = \bA(t) \bP(t) + \bP(t) \bA^T(t) + \bQ(t) \label{eq:OCPdP} \\
& \phantom{\text{s.t.} \quad} \bP(0)  = \bP_0 \succeq 0 \label{eq:OCPdPIC} \\ %& \phantom{\text{s.t.} \qquad} 0       \geq h_i(\bmu(t), \bu(t))
%+ \overbrace{\gamma \bigg[\underbrace{\na^T_{\bx} h_i \bP(t) \na_{\bx} h_i}_{\text{variance of $h_i(\bx)$}}\bigg]^{\frac{1}{2}}}^{\text{safety margin}},
&\phantom{\text{s.t.} \qquad} 0       \geq h_i(\bmu(t), \bu(t))
+ \be_i(\bmu(t), \bu(t), \bP(t)),
\quad i \in \calI \label{eq:OCPconstraints}
\end{align}
The cost function $J$ in~\eqref{eq:OCPcost} depends on the mean $\bmu(t)$, the controls $\bu(t)$ and the state covariance $\bP(t)$. The mean dynamics is expressed by~\eqref{eq:OCPdyn} with~\eqref{eq:OCPdynIC}, and the covariance dynamics is expressed by~\eqref{eq:OCPdP} with~\eqref{eq:OCPdPIC}. In eq.~\eqref{eq:OCPconstraints} the \emph{backoff terms} $\be_i$ account for the disturbances and serve the purpose of obtaining deterministic safety-margin types of constraint directly on $\bmu(t)$. $\calI$ is the set of indices defining the inequality constraints. This is equivalent to a linearization around the mean of the original \emph{chance constraint} on $\bx(t)$ expressed by $\Prob \{h_i(\bx) \leq 0\}\geq p$, where $p$ is the confidence level of constraint satisfaction. The backoff terms are given by
$\be_i = \gamma \sig_i$. The coefficient $\gamma = \Phi^{-1}(p)$ is the quantile function, where $\Phi(z)=\Prob \{Z\leq z\}$ is the cdf of a standard normal distribution $Z \sim \calN(0,1)$, and acts as a \emph{tuning knob}: the greater the confidence level $p$ required, the higher the gain $\ga$\footnote{For example, with $p=0.84$ $\ga = 1.0$, with $p=0.97$ $\ga = 2.0$, with $p=0.99$ $\ga = 3.0$}. The term $\sig_i = \big[\na^T_{\bx} h_i(\bmu) \bP(t) \na_{\bx} h_i(\bmu)\big]^{\frac{1}{2}}$ represents the standard deviation of the constraint $h_i(\bx)$ linearized around the mean, i.e. of random variable $h_i(\bmu)+\na^T_{\bx} h_i(\bmu) (\bx - \bmu)$, and follows immediately from the propagation rule of covariance.

@@

In certain cases~\cite{Gillis:PracticalMethodsApproximate:2015}, positive-definiteness-preserving Lyapunov discretization schemes are used in~\eqref{eq:DOCPdP}. However, for sufficiently fine discretization, we found that integrating only the lower triangular part of $\bP(t)$ in~\eqref{eq:OCPdPIC}  (and reconstructing its strictly upper part accordingly), ensures that all $\bP_k$ remain symmetric and positive definite when the initial $\bP_0$ is symmetric and positive definite. In all our tests we used direct collocation with a cubic polynomial and Gauss-Legendre collocation points.
\subsection{External Wrenches}
\label{sec:extwren}
The external wrenches acting on each body, from wheels to chassis, are expressed in body-fixed components. Only gravity, aerodynamic forces, and road-tyre interactions are considered. Following~\cite{Guiggiani:book:2023}, the aerodynamic wrench in frame $\{B\}$ is $W_a=-\frac{1}{2}\rho S v_{gb,1}^2 [C_x, 0, C_z, 0, -h_0 C_x - a_f C_{zf} + a_r C_{zr}, 0]^T$, where $\rho$ is air density, $S$ is the frontal area, $v_{gb,1}$ the forward velocity (component of \emph{distal rigid-body twist} $V_{gb}$, see~\cite[p.~4]{Domenighini:Designs:2023}), and $C_x, C_z$ are drag/lift coefficients. $C_z$ is split into $C_{zf}=k_a C_z$ and $C_{zr}=(1-k_a)C_z$ by the aerodynamic balance coefficient $k_a$.

>>>>

Accordingly, the continuous-time stochastic trajectory planning can be framed as the following nonlinear optimal control problem
\begin{align} 
& \underset{\bmu(t), \bu(t), \bP(t)}{\text{minimize}} \quad J(\bmu(t), \bu(t), \bP(t)) \label{eq:OCPcost} \\
& \text{s.t.} \quad \dbmu(t)           = f(\bmu(t), \bu(t)) \label{eq:OCPdyn} \\
& \phantom{\text{s.t.} \quad} \bmu(0)  = \bmu_0 \label{eq:OCPdynIC} \\
& \phantom{\text{s.t.} \quad} \dbP(t) = \bA(t) \bP(t) + \bP(t) \bA^T(t) + \bQ(t) \label{eq:OCPdP} \\
& \phantom{\text{s.t.} \quad} \bP(0)  = \bP_0 \succeq 0 \label{eq:OCPdPIC} \\ %& \phantom{\text{s.t.} \qquad} 0       \geq h_i(\bmu(t), \bu(t))
%+ \overbrace{\gamma \bigg[\underbrace{\na^T_{\bx} h_i \bP(t) \na_{\bx} h_i}_{\text{variance of $h_i(\bx)$}}\bigg]^{\frac{1}{2}}}^{\text{safety margin}},
&\phantom{\text{s.t.} \qquad} 0       \geq h_i(\bmu(t), \bu(t))
+ \be_i(\bmu(t), \bu(t), \bP(t)),
\quad i \in \calI \label{eq:OCPconstraints}
\end{align}
The cost function $J$ in~\eqref{eq:OCPcost} depends on the mean $\bmu(t)$, the controls $\bu(t)$ and the state covariance $\bP(t)$. The mean dynamics is expressed by~\eqref{eq:OCPdyn} with~\eqref{eq:OCPdynIC}, and the covariance dynamics is expressed by~\eqref{eq:OCPdP} with~\eqref{eq:OCPdPIC}. In eq.~\eqref{eq:OCPconstraints} the \emph{backoff terms} $\be_i$ account for the disturbances and serve the purpose of obtaining deterministic safety-margins directly on $\bmu(t)$. $\calI$ is the set of indices defining the inequality constraints. This is equivalent to a linearization around the mean of the original \emph{chance constraint} on $\bx(t)$ expressed by $\Prob \{h_i(\bx) \leq 0\}\geq p$, where $p$ is the confidence level of constraint satisfaction. The backoff terms are given by
$\be_i = \gamma \sig_i$. The coefficient $\gamma = \Phi^{-1}(p)$ is the quantile function, where $\Phi(z)=\Prob \{Z\leq z\}$ is the cdf of a standard normal distribution $Z \sim \calN(0,1)$, and acts as a \emph{tuning knob}: the greater the confidence level $p$ required, the higher the gain $\ga$\footnote{For example, with $p=0.84$ $\ga = 1.0$, with $p=0.97$ $\ga = 2.0$, with $p=0.99$ $\ga = 3.0$}. The term $\sig_i = \big[\na^T_{\bx} h_i(\bmu) \bP(t) \na_{\bx} h_i(\bmu)\big]^{\frac{1}{2}}$ represents the standard deviation of the constraint $h_i(\bx)$ linearized around the mean, i.e. of random variable $h_i(\bmu)+\na^T_{\bx} h_i(\bmu) (\bx - \bmu)$, and follows immediately from the propagation rule of covariance.

@@

In certain cases~\cite{Gillis:PracticalMethodsApproximate:2015}, positive-definiteness-preserving Lyapunov discretization schemes are used in~\eqref{eq:DOCPdP}. However, for sufficiently fine discretization, we found that integrating only the lower triangular part of $\bP(t)$ in~\eqref{eq:OCPdPIC}  (and reconstructing its strictly upper part accordingly), ensures that all $\bP_k$ remain symmetric and positive definite when the initial $\bP_0$ is symmetric and positive definite. In all our tests direct collocation with cubic polynomial state representations and Gauss-Legendre collocation points were used.
\subsection{External Wrenches}
\label{sec:extwren}
The external wrenches acting on each body, from wheels to chassis, are expressed in body-fixed components. Only gravity, aerodynamic forces, and road-tyre interactions are considered. Following~\cite{Guiggiani:book:2023}, the aerodynamic wrench in frame $\{B\}$ is $W_a=-\frac{1}{2}\rho S v_{gb,1}^2 [C_x, 0, C_z, 0, -h_0 C_x - a_f C_{zf} + a_r C_{zr}, 0]^T$, where $\rho$ is air density, $S$ is the frontal area, $v_{gb,1}$ the forward velocity (component of \emph{distal rigid-body twist} $V_{gb}$, see~\cite[p.~4]{Domenighini:Designs:2023}), and $C_x, C_z$ are drag/lift coefficients. $C_z$ is split into $C_{zf}=k_a C_z$ and $C_{zr}=(1-k_a)C_z$ by the aerodynamic balance coefficient $k_a$.
